Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-05-04 inspiral_svd.tex
added flop budget
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@740 ccdd05d5-56c1-461a-bf98-85f6876523b4

\begin{abstract}
The rapid detection of compact binary coalescence with a network of advanced
gravitational-wave detectors offers a unique opportunity for multimessenger
Astronomy.  Prompt alerts to the Astronomical community will assure that all
available instruments are observing the electromagnetic counterparts to the
first gravitational waves from coalescing compact objects.   Rapid, near
realtime alerts, will improve the chances of detecting weak, decaying
electromagnetic signals and correlating such signals with the gravitational
wave strain.  In this paper, we describe a new, realtime, detection strategy
that exploits the properties of inspiral waveforms using principal component
analysis and multi-rate filtering.  We provide detection efficiency
characteristics from a prototype search pipeline that is capable of
\earlywarning\ detection with acheivable computational requirements.  
\end{abstract}

>>>>

\begin{abstract}
The rapid detection of compact binary coalescence with a network of advanced
gravitational-wave detectors offers a unique opportunity for multimessenger
Astronomy.  Prompt alerts to the Astronomical community will assure that all
available instruments are observing the electromagnetic counterparts to the
first gravitational waves from coalescing compact objects.   Rapid, near
realtime alerts, will improve the chances of detecting weak, decaying
electromagnetic signals and correlating such signals with the gravitational
wave strain.  In this paper, we describe a detection strategy that exploits the
properties of inspiral waveforms using principal component analysis and
multi-rate filtering.  We provide detection efficiency characteristics from a
prototype search pipeline that is capable of \earlywarning\ detection with
acheivable computational requirements.  
\end{abstract}

### 2011-05-04 introduction.tex
added flop budget
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@740 ccdd05d5-56c1-461a-bf98-85f6876523b4

In 2010, \textsc{ligo} completed its sixth science run (S6) and Virgo
completed its third science run (VSR3). When both \textsc{ligo} and Virgo were
operating, several all-sky detection pipelines operated in a low-latency
configuration~\cite{HugheyGWPAW2011, S6lowlatency}. Candidates were sent for
electromagnetic followup to several telescopes including Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008, HugheyGWPAW2011}.
The S6 analyses achieved latencies of $~\sim30$--$60$ minutes, which were
dominated by a human vetting process. The intrinsic gravitational-wave trigger
generator latency was $\sim$2--5 minutes.  In this work we will assume that
the human vetting process can be automated in the advanced detector era and
that the dominating latency will be trigger generation. From this perspective
we will try to address the realistic expectations for reducing trigger latency
from $\sim$ minutes \emph{after} the merger to minutes \emph{before} the
merger for sufficiently loud events.  Figure \ref{fig:earlywarning} presents
the possibility for early detection of NS--NS binaries with Advanced \LIGO\
assuming the rate predictions in \cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning}\FIXME{Drew has independently reproduced
this calculation} Expected number of NS--NS sources that will be detectable
$T_{bc}$ seconds before coalescence.  The solid line is the most likely yearly
rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the shaded region is
the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within
10 seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early detection
but at the cost of fewer total events observed above SNR 8.}
\end{figure}

>>>>

In 2010, \textsc{ligo} completed its sixth science run (S6) and Virgo
completed its third science run (VSR3). When both \textsc{ligo} and Virgo were
operating, several all-sky detection pipelines operated in a low-latency
configuration~\cite{HugheyGWPAW2011, S6lowlatency}. Candidates were sent for
electromagnetic followup to several telescopes including Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008, HugheyGWPAW2011}.
The S6 analyses achieved latencies of $~\sim30$--$60$ minutes, which were
dominated by a human vetting process. The intrinsic gravitational-wave trigger
generator latency was $\sim$2--5 minutes.  In this work we will assume that
the human vetting process can be automated in the advanced detector era and
that the dominating latency will be trigger generation. From this perspective
we will try to address the realistic expectations for reducing trigger latency
from $\sim$ minutes \emph{after} the merger to minutes \emph{before} the
merger for sufficiently loud events.  Figure \ref{fig:earlywarning} presents
the possibility for early detection of NS--NS binaries with Advanced \LIGO\
assuming the rate predictions in \cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}
